---
title: "CAS Common Chemistry API in R"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  fig.path = "_figures/casc/"
)
```

# CAS Common Chemistry API in R

by Adam M. Nguyen and Michael T. Moen

<div class="rmd-btn-wrapper">
  <a class="rmd-btn"
     href="https://github.com/UA-Libraries-Research-Data-Services/UALIB_ScholarlyAPI_Cookbook/blob/main/rmarkdown/casc.Rmd"
     target="_blank"
     rel="noreferrer">
    View RMarkdown File
  </a>
</div>

The CAS Common Chemistry API provides programmatic access to a curated subset, offering information on nearly 500,000 chemical substances.

Please see the following resources for more information on API usage:

- Documentation
    - <a href="https://commonchemistry.cas.org/" target="_blank">CAS Common Chemistry</a>
    - <a href="https://commonchemistry.cas.org/api-overview" target="_blank">CAS Common Chemistry API Overview</a>
- Terms
    - <a href="https://www.cas.org/legal" target="_blank">CAS Policies</a>
- Data Reuse
    - CAS Common Chemistry is provided under the <a href="https://creativecommons.org/licenses/by-nc/4.0/" target="_blank">Creative Commons CC BY-NC 4.0 license</a>

**_NOTE:_** Please see access details and rate limit requests for this API in the official documentation.

*These recipe examples were tested on March 24, 2026.*

## Setup

### Load Libraries

The following packages need to be installed into your environment to run the code examples in this tutorial. These packages can be installed with `install.packages()`.

- <a href="https://cran.r-project.org/web/packages/httr/index.html" target="_blank">httr: Tools for Working with URLs and HTTP</a>
- <a href="https://cran.r-project.org/web/packages/jsonlite/index.html" target="_blank">jsonlite: A Simple and Robust JSON Parser and Generator for R</a>
- <a href="https://cran.r-project.org/web/packages/magick/index.html" target="_blank">magick: Advanced Graphics and Image-Processing in R</a>

We load the libraries used in this tutorial below:

```{r library, warning=FALSE, message=FALSE}
library(httr)
library(jsonlite)
library(magick)
```

### Import API Key

An API key is required to access the CAS Common Chemistry API. You can sign up for one at the <a href="https://www.cas.org/services/commonchemistry-api" target="_blank">Request API Access for CAS Common Chemistry</a> page.

We keep our token in a `.Renviron` file that is stored in the working directory and use `Sys.getenv()` to access it. The `.Renviron` should have an entry like the one below.

```text
CAS_API_KEY="PUT_YOUR_API_KEY_HERE"
```

Below, we can test to whether the key was successfully imported.

```{r}
if (nzchar(Sys.getenv("CAS_API_KEY"))) {
  print("API key successfully loaded.")
} else {
  warning("API key not found or is empty.")
}
```

## 1. Common Chemistry Record Detail Retrieval

Common Chemistry is an extremely useful open community resource for accessing chemical information for nearly 500,000 chemical chemistry from the CAS registry. In this example, we detail the steps for accessing a simple example chemical substance, ethyl cyclohexanepropionate.

### API Parameter Setup   

First, let's setup the API parameters for Common Chemistry. Below we are going to provide a variable `detail_base_url` to query Common Chemistry for details and more specifically. In this example, using a registry number, or as we abbreviate it an `rn`.

```{r, warning=FALSE, message = FALSE}
DETAIL_BASE_URL <- "https://commonchemistry.cas.org/api/detail?"
casrn1 <- "10094-36-7" # ethyl cyclohexanepropionate
```

### Request Data from CAS Common Chemistry Detail API

Using the `httr` and `jsonlite` libraries, we query the API and convert the data to a usable format.

```{r, warning=FALSE, message=FALSE}
url <- paste0(DETAIL_BASE_URL, "cas_rn=", casrn1)

# Use httr's 'GET()' function to retrieve raw data from a URL
casrn1_response <-GET(url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
casrn1_response
```

The `GET()` function returns `Response`, `Date`, `Status: 200` indicating a successful request, `Size`, and `Content-Type`. In most cases when accessing APIs, the `Content-Type` returns JSON, so we need the `jsonlite` to parse it.

```{r convert}
# Extract JSON data from the response
casrn1_data <- fromJSON(rawToChar(casrn1_response$content))

casrn1_data
```

### Accessing Specific Data

The `$` operator can be used to extract specific parts of `casrn1_data`.

```{r warning=FALSE, message=FALSE}
# Access the canonical SMILE
casrn1_smile <- casrn1_data$canonicalSmile
casrn1_smile
```

```{r warning=FALSE, message=FALSE}
# Another example may be finding synonyms for Ethyl cyclohexanepropionate
casrn1_synonyms <- casrn1_data$synonyms
casrn1_synonyms
```

The methods above work for any parts of the data object, to find a list of the data objects, you can use the `ls()` function.

```{r, warning=FALSE, message=FALSE}
ls(casrn1_data)
```

Given this list we can select different data as seen in previous examples, i.e. `casrn1_data$(example)`.

### Display the Molecule Drawing      

The `images` section of the data contains data formatted as Scalable Vector Graphics (SVG). SVG is an XML based vector format for defining two-dimensional graphics and has been an open standard since 1999. Using the SVG data provided by the API, we display the molecule drawing using the package `magick`, a useful toolkit for image processing in R.

```{r molecule-drawing, warning=FALSE, message=FALSE}
image_str <- charToRaw(casrn1_data$image)
image_read(image_str, density = 300)
```

## 2. Common Chemistry API Record Detail Retrieval in a Loop   

In this example, we show how to query Common Chemistry for details given a list of registry numbers.

```{r, warning=FALSE, message=FALSE}
# Create a list of CAS RNs
casrns <- c(
  "10094-36-7",
  "10031-92-2",
  "10199-61-8",
  "10036-21-2",
  "1019020-13-3"
)

# Iterate through list of CAS RN and append to list
list <- lapply(casrns, function(casrn) {
  url <- paste0(DETAIL_BASE_URL, "cas_rn=", casrn)
  response <- GET(url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
  casrn_data <- fromJSON(rawToChar(response$content))
  Sys.sleep(1)
  return(casrn_data)
})

# Show first result in the list
head(list, n = 1)
```

We now have a list of sub-lists of data corresponding to the CAS RNs.

### Display Images

```{r molecule-loop}
# Use 'lapply' to iterate through list and convert our image data to a readable format
imgs <- lapply(list, function(x) {charToRaw(x$image)})

# Use 'image_read()' to read SVGs

# "Ethyl cyclohexanepropionate"
image_read(imgs[[1]], density = 300)
# "Ethyl 2-nonynoate"
image_read(imgs[[2]], density = 300)
# "Ethyl 1<em>H</em>-pyrazole-1-acetate"
image_read(imgs[[3]], density = 300)
# "Ethyl 3-(ethoxycarbonyl)benzenepropanoate"
image_read(imgs[[4]], density = 300)
# "Ethyl 1-cyclohexene-1-carboximidate"
image_read(imgs[[5]], density = 300)
```

### Select Specific Data    

We use `lapply()` to iterate through the list and grab each sub-list's `canonicalSmile`.

```{r specdat, warning=FALSE, message=FALSE}
# Apply function(i) to receive canonicalSmiles for every instance in the list
cansmiles <- lapply(list, function(i) {i$canonicalSmile})
unlist(cansmiles) # Flatten list
```

Through similar means, we can also iterate through the list and grab each sub-list's synonyms.

```{r getsyns, warning=FALSE, message =FALSE}
synonyms_list <- lapply(list, function(j) {j$synonyms})
synonyms_list
```

```{r flatsyns, warning=FALSE, message =FALSE}
# Use the unlist() function to flatten our list of synonyms
flatsyns <- unlist(synonyms_list)
flatsyns
```

### Create Data Frame   

Let's create a data frame from `list` with the variables `uri`, `rn`, `name`, `inchiKey`, `canonicalSmile`, and `molecularMass`.

```{r list2df, warning=FALSE, message=FALSE}
df <- data.frame(do.call
        # Iterate through list of chemicals with function(i) and append to df
        (rbind, lapply(list, function(item)
          cbind(
            item["uri"],
            item["rn"],
            item["name"],
            item["inchiKey"],
            item["canonicalSmile"],
            item["molecularMass"]
          ))),
        row.names = NULL)

names(df) <- c("uri", "rn", "name", "inchiKey", "canonicalSmile", "molecularMass")

# Display df
df
```

## 3. Common Chemistry 

In addition to the `/detail` endpoint, the CAS Common Chemistry API has a `/search` endpoint that allows searching by CAS RN, SMILES, InChI/InChIKey, and name.

### Request data form CAS Common Chemistry Search API

```{r search, warning=FALSE, message = FALSE}
SEARCH_BASE_URL <- "https://commonchemistry.cas.org/api/search?q="

# InChIKey for Quinine
IK <- "InChIKey=LOUPRKONTZGTKE-WZBLMQSHSA-N"
url <- paste0(SEARCH_BASE_URL, IK)
search_response <- GET(url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
search_response
```

Like before, we need to access the `content` part of the data to retrieve what we are interested. This data is stored in JSON.

```{r, warning=FALSE,  message=FALSE}
# Extract JSON data from response
quinine_rn <- fromJSON(rawToChar(search_response$content))

quinine_rn
```

Now we have the quinine's CAS RN, so we can query quinine based on this, similar to the first example.

```{r warning=FALSE, message=FALSE}
# Get detailed record for quinine
url <- paste0(DETAIL_BASE_URL, "cas_rn=", quinine_rn$results$rn)
detail_response <- GET(url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
detail_response
```

```{r warning=FALSE, message=FALSE}
# Convert JSON structure to character format
quinine_data <- fromJSON(rawToChar(detail_response$content))

quinine_data
```

### Handle Multiple Results

```{r multiple_results, warning=FALSE, message=FALSE}
# SMILES for butadiene
smi_bd <- "C=CC=C"

# Request data from CAS Common Chemistry Search API
url <- paste0(SEARCH_BASE_URL, smi_bd)
search_response <- GET(url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))

smi_search_data <- fromJSON(rawToChar(search_response$content))

# Get results count
smi_search_data$count

# Display CAS RNs
smi_search_data$results$rn

names <- lapply(smi_search_data$results$rn, function(casrn) {
  
  smi_url <- paste0(DETAIL_BASE_URL, "cas_rn=", casrn)
  smi_data <- GET(smi_url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
  smi_data <- fromJSON(rawToChar(smi_data$content))
  
  Sys.sleep(1)
  return(smi_data$name)
})
unlist(names) # Display names
```

### Handle Multiple Page Results   

The CAS Common Chemistry API returns 50 results per page, and only the first page is returned by default. If the search returns more than 50 results, the offset option can be added to page through and obtain all results:

```{r multi_page, warning=FALSE, message= FALSE}
# Set up search query parameters
n <- "selen*"

# Calculate number of pages needed
num_results <- 200  # Total number of results to retrieve
page_size <- 50     # Number of results to retrieve per page
num_pages <- ceiling(num_results / page_size)

# Create empty list for results
n_search_data <- list()

# Loop through the pages of data
for (page_idx in seq_len(num_pages)) {
  # Calculate the starting offset for the current page
  offset <- (page_idx - 1) * page_size
  
  # Construct the search URL for the current page
  search_url <- paste0(SEARCH_BASE_URL, n, "&offset=", offset)
  
  # Make the API request and append the response to the list of results
  response <- GET(search_url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
  n_search_data[[page_idx]] <- fromJSON(rawToChar(response$content))
  
  # Pause for 1 second
  Sys.sleep(1)
}

# Length of search data includes a top level list for each query
length(n_search_data)

# List within lists contain the results
length(n_search_data[[1]]$results)
length(n_search_data[[2]]$results)
length(n_search_data[[3]]$results)
length(n_search_data[[4]]$results)

# We can index and extract out the first CAS RN as follows
n_search_data[[1]]$results[[1]][[1]]
```

```{r, warning=FALSE, message=FALSE}
n_casrn_list <- character()
for (n_idx in seq_along(n_search_data)) {
  n_casrn_list <- c(n_casrn_list, n_search_data[[n_idx]]$results$rn)
}
length(n_casrn_list)
```

```{r}
n_casrn_list[1:10]
```

We can now iterate through the list of CAS RNs to store data about the chemicals. In this example we extract a list of `molecularMass` values and visualize them.

```{r, warning=FALSE, message=FALSE}
# Now we can loop through each CAS RN and use the detail API to obtain the entire record
# This will query CAS Common Chem 191 times and take ~ 5 min.
DETAIL_BASE_URL <- "https://commonchemistry.cas.org/api/detail?"
n_detail_data <- list()
for (casrn in n_casrn_list) {
  casrn_url <- paste0(DETAIL_BASE_URL, "cas_rn=", casrn)
  response <- GET(casrn_url, add_headers("X-API-KEY" = Sys.getenv("CAS_API_KEY")))
  n_detail_data[[casrn]] <- content(response)
  Sys.sleep(1) # Add a 1 second delay between API calls
}
```

```{r moleMassCont., warning=FALSE, message=FALSE}
mms <- list()
for (mm_idx in seq_along(n_detail_data)) {
  mms <- c(mms, n_detail_data[[mm_idx]]$molecularMass)
}
head(mms, n = 10)
```

### Visualization   

Using the list of molecular mass values we collected, we can create visualizations in R using the `hist()` and `density()` functions.

```{r molecular-mass-histogram, warning=FALSE, message=FALSE}
# Histogram
hist(as.numeric(mms), # Reformat mms type to numeric and call as x variable
     breaks = 20,
     main = "Histogram of available molecularMass values", # Set title
     xlab = "Molecular Mass", # X Label
     ylab = "Count", # Y Label
     col = "plum", ) # Color
```

We can also make a kernel density plot, a graph that uses a continuous curve to depict the distribution of the data with respect to the molecular mass values. This will give some more fidelity as opposed to the histogram.

```{r molecular-mass-kernel, warning=FALSE, message=FALSE}
d <- density(na.omit(as.numeric(mms)))
plot(d, main = "Kernel Density of Available mms values")
polygon(d, col = "plum", border = "black")
```
