---
title: "College Scorecard API in R"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "_figures/college-scorecard/"
)
```

# College Scorecard API in R

by Michael T. Moen

<div class="rmd-btn-wrapper">
  <a class="rmd-btn"
     href="https://github.com/UA-Libraries-Research-Data-Services/UALIB_ScholarlyAPI_Cookbook/blob/main/rmarkdown/college-scorecard.Rmd"
     target="_blank"
     rel="noreferrer">
    View RMarkdown File
  </a>
</div>

The College Scorecard API is an online tool hosted by the U.S. Department of Education that contains data concerning higher education institutions.

Please see the following resources for more information on API usage:

- Documentation
    - <a href="https://collegescorecard.ed.gov/data/api-documentation/" target="_blank">College Scorecard API Documentation</a>
- Data Reuse
    - <a href="https://www.ed.gov/about/ed-overview/required-notices/website-policies/copyright-status-notice-us-department-of-education" target="_blank">College Scorecard Copyright Status</a>
    - <a href="https://data.gov/privacy-policy/" target="_blank">Data\.gov Privacy Policy</a>

**_NOTE:_** Please see access details and rate limit requests for this API in the official documentation.

*These recipe examples were tested on March 23, 2026.*

## Setup

### Load Libraries

The following packages need to be installed into your environment to run the code examples in this tutorial. These packages can be installed with `install.packages()`.

- <a href="https://cran.r-project.org/web/packages/httr/index.html" target="_blank">httr: Tools for Working with URLs and HTTP</a>
- <a href="https://cran.r-project.org/web/packages/jsonlite/index.html" target="_blank">jsonlite: A Simple and Robust JSON Parser and Generator for R</a>

We load the libraries used in this tutorial below:

```{r import-libraries}
library(httr)
library(jsonlite)
```

### Import API Key

An API key is required for to access the College Scorecard API. You can sign up for one at <a href="https://collegescorecard.ed.gov/data/api-documentation#api-access-and-authentication" target="_blank">College Scorecard Website</a>.

We keep our token in a `.Renviron` file that is stored in the working directory and use `Sys.getenv()` to access it. The `.Renviron` should have an entry like the one below.

```text
COLLEGE_SCORECARD_API_KEY="PUT_YOUR_API_KEY_HERE"
```

Below, we can test to whether the key was successfully imported.

```{r import-api-key}
if (nzchar(Sys.getenv("COLLEGE_SCORECARD_API_KEY"))) {
  print("API key successfully loaded.")
} else {
  warning("API key not found or is empty.")
}
```

## 1. Get Names of All Institutions

To start, we'll use a basic query to find the names of all educational institutions recognized by the College Scorecard API.

All of the data for the API can be found using the `v1/schools` endpoint.

Fields in the College Scorecard API are accessed with a `<time>.<category>.<name>` sequence:
- `<time>` indicates the year of the data to be accessed. To access the most recent data, use `latest`.
- `<category>` and `<name>` can be found in the Data Dictionary file that can be downloaded from the API's documentation. The `<category>` of a field is given by the `dev-category` column in the `Institution_Data_Dictionary` section, and the `<name>` is given by the `developer-friendly name` column.

```{r}
# Define base URL
BASE_URL <- "https://api.data.gov/ed/collegescorecard/v1/schools?"

# Define parameters
params <- list(
  fields = "school.name",
  api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
)

# Perform HTTP GET request
response <- GET(BASE_URL, query = params)

# Status code 200 indicates success
response$status_code
```

```{r}
names_data <- fromJSON(rawToChar(response$content))
names_data$metadata
```

The `total` value indicates the total number results returned in this query. These results are paginated, so each query will return only the number indicated by `page_size`, which has a default value of 20 and a maximum value of 100. The page number is indicated by `page`, which by default is set to 0.

We can use a loop to create an API request for each page:

```{r}
field <- "school.name"
sort_key <- "school.name"
page_size <- 100

total_pages <- ceiling(names_data$metadata$total / page_size)
institution_names <- c()

for (page_number in 0:(total_pages - 1)) {
  params <- list(
    fields = field,
    page = page_number,
    per_page = page_size,
    sort = sort_key,
    api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
  )
  response <- GET(BASE_URL, query = params)
  name_data <- fromJSON(rawToChar(response$content))$results
  institution_names <- c(institution_names, name_data$school.name)
  Sys.sleep(1)
}

# Print number of institutions returned
length(institution_names)
```

```{r}
# Print first 10 institution names
institution_names[1:10]
```

## 2. Get Names of All Universities

College Scorecard API requests can also take conditions to only select certain institutions.

In this example, we limit the results to only include institutions that award graduate degrees. In order to do this, we set the `degrees_awarded.highest` parameter to `4` to indicate that the highest degree awarded by an institution is a graduate degree. This information is within the `Institution_Data_Dictionary` section of the College Scorecard data dictionary.

```{r}
page_size <- 100
field <- "school.name"
sort_key <- "school.name"

params <- list(
  fields = field,
  latest.school.degrees_awarded.highest = 4,
  api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
)

response <- GET(BASE_URL, query = params)
names_data <- fromJSON(rawToChar(response$content))
total_pages <- ceiling(names_data$metadata$total / page_size)

university_names <- c()
for (page_number in 0:(total_pages - 1)) {
  params <- list(
    fields = field,
    latest.school.degrees_awarded.highest = 4,
    sort = sort_key,
    page = page_number,
    per_page = page_size,
    api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
  )
  response <- GET(BASE_URL, query = params)
  name_data <- fromJSON(rawToChar(response$content))$results
  university_names <- c(university_names, name_data[[field]])
  Sys.sleep(1)
}

# Print number of university names found
length(university_names)
```

```{r}
# Print first 10 university names
university_names[1:10]
```

## 3. Find Number of Universities by State

The `school.state_fips` data element contains a number that corresponds to each state. This mapping is given below:

```{r}
states <- list(
  "1" = "Alabama", "2" = "Alaska", "4" = "Arizona", "5" = "Arkansas", "6" = "California",
  "8" = "Colorado", "9" = "Connecticut", "10" = "Delaware", "11" = "District of Columbia",
  "12" = "Florida", "13" = "Georgia", "15" = "Hawaii", "16" = "Idaho", "17" = "Illinois",
  "18" = "Indiana", "19" = "Iowa", "20" = "Kansas", "21" = "Kentucky", "22" = "Louisiana",
  "23" = "Maine", "24" = "Maryland", "25" = "Massachusetts", "26" = "Michigan",
  "27" = "Minnesota", "28" = "Mississippi", "29" = "Missouri", "30" = "Montana",
  "31" = "Nebraska", "32" = "Nevada", "33" = "New Hampshire", "34" = "New Jersey",
  "35" = "New Mexico", "36" = "New York", "37" = "North Carolina", "38" = "North Dakota",
  "39" = "Ohio", "40" = "Oklahoma", "41" = "Oregon", "42" = "Pennsylvania",
  "44" = "Rhode Island", "45" = "South Carolina", "46" = "South Dakota",  "47" = "Tennessee",
  "48" = "Texas", "49" = "Utah", "50" = "Vermont", "51" = "Virginia", "53" = "Washington",
  "54" = "West Virginia", "55" = "Wisconsin", "56" = "Wyoming", "60" = "American Samoa",
  "64" = "Federated States of Micronesia", "66" = "Guam", "69" = "Northern Mariana Islands",
  "70" = "Palau", "72" = "Puerto Rico", "78" = "Virgin Islands"
)
```

Using this mapping, we can find the number of universities in each state:

```{r}
field <- "latest.school.state_fips"
page_size <- 100

# Calculate the number of loops needed to page through every result
params <- list(
  latest.school.degrees_awarded.highest = 4,
  fields = field,
  api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
)
response <- GET(BASE_URL, query = params)
state_data <- fromJSON(rawToChar(response$content))
total_pages <- ceiling(names_data$metadata$total / page_size)

state_freq <- list()
for (page_number in 0:(total_pages - 1)) {

  params <- list(
    latest.school.degrees_awarded.highest = 4,
    fields = field,
    page = page_number,
    per_page = page_size,
    api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
  )
  response <- GET(BASE_URL, query = params)
  state_data <- fromJSON(rawToChar(response$content))$results

  state_fips_codes <- as.character(state_data$latest.school.state_fips)
  for (state_fips in state_fips_codes) {
    state_name <- states[[state_fips]]
      state_freq[[state_name]] <- ifelse(is.null(state_freq[[state_name]]), 1, state_freq[[state_name]] + 1)
  }
  Sys.sleep(1)
}
```

Now, we can sort and display the results:

```{r}
# Print the top 20 states/territories with the most universities
head(sort(unlist(state_freq), decreasing = TRUE), 20)
```

## 4. Retrieve Multiple Data Points in a Single Query

The following example uses multiple conditions and multiple fields. The conditions in the query are separated by `&` while the fields are separated by `,`.

```{r}
# Join fields into a comma-delimited string
fields <- paste(
  "school.name",
  "latest.admissions.admission_rate.overall",
  "latest.student.size",
  "latest.cost.tuition.out_of_state",
  "latest.cost.tuition.in_state",
  "latest.student.demographics.median_hh_income",
  "latest.school.endowment.begin",
  sep = ","
)
sort_key <- "school.name"
page_size <- 100

# Calculate the number of loops needed to page through every result
params <- list(
  fields = fields,
  latest.school.degrees_awarded.highest = 4,
  latest.student.size__range = "1000..",  # Schools with 1000 or more students
  api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
)

response <- GET(BASE_URL, query = params)
names_data <- fromJSON(rawToChar(response$content))
total_pages <- ceiling(names_data$metadata$total / page_size)

rows <- list()
for (page_number in 0:(total_pages - 1)) {
  params <- list(
    fields = fields,
    latest.school.degrees_awarded.highest = 4,
    latest.student.size__range = "1000..",
    page = page_number,
    per_page = page_size,
    sort = sort_key,
    api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
  )
  response <- GET(BASE_URL, query = params)
  data <- fromJSON(rawToChar(response$content))$results
  for (university in 1:nrow(data)) {
    row <- list(
      Name = data$school.name[university],
      Admission_Rate = data$latest.admissions.admission_rate.overall[university],
      Size = data$latest.student.size[university],
      Tuition_Out_State = data$latest.cost.tuition.out_of_state[university],
      Tuition_In_State = data$latest.cost.tuition.in_state[university],
      Median_HH_Income = data$latest.student.demographics.median_hh_income[university],
      Endowment = data$latest.school.endowment.begin[university]
    )
    rows <- append(rows, list(row))
  }
  Sys.sleep(1)
}

df <- do.call(rbind.data.frame, rows)

# Print first 10 rows
head(df, 10)
```

We can query the resulting data frame to find the data for specific universities:

```{r}
ua_df <- df[df$Name == "The University of Alabama", ]
print(ua_df)
```

We can also query the data frame to find the data for universities that satisfy certain conditions:

```{r}
filtered_df <- df[df$Admission_Rate < 0.1, ]
filtered_df <- na.omit(filtered_df)
filtered_df <- filtered_df[order(filtered_df$Admission_Rate), ]
filtered_df
```

```{r}
filtered_df <- df[df$Endowment > 1.0e+10, ]
filtered_df <- na.omit(filtered_df)
filtered_df <- filtered_df[order(-filtered_df$Endowment), ]
filtered_df
```

## 5. Retrieve All Data for an Institution

The College Scorecard API can also be used to retrieve all of the data for a particular institution. The example below finds all data for The University of Alabama:

```{r}
params <- list(
  school.name = "The University of Alabama",
  api_key = Sys.getenv("COLLEGE_SCORECARD_API_KEY")
)
response <- GET(BASE_URL, query = params)
ua_data <- fromJSON(rawToChar(response$content))$results

# Print structure of the result
str(ua_data, max.level = 1)
```

Finally, we'll look at the breakdown of size of each program at the University of Alabama:

```{r ua-program-pie-chart}
program_percentage_data <- ua_data[[1]]$academics$program_percentage
perc <- unlist(program_percentage_data)

threshold <- 0.03
small <- perc < threshold

if (sum(small) > 1) {
  perc <- c(perc[!small], other = sum(perc[small]))
}

par(mar = c(5, 4, 4, 8))

pie(
  perc,
  labels = paste0(round(perc * 100, 1), "%"),
  col = rainbow(length(perc)),
  main = "Program Percentage at The University of Alabama"
)

legend(
  "topright",
  inset = c(-0.25, 0),
  xpd = TRUE,
  legend = names(perc),
  fill = rainbow(length(perc)),
  cex = 0.8
)
```

```{r}
# Sort the list by values in descending order
sorted_program_percentage_data <- program_percentage_data[order(-unlist(program_percentage_data))]

# Print the sorted data
for (key in names(sorted_program_percentage_data)) {
  cat(paste(key, ":", sorted_program_percentage_data[[key]], "\n"))
}
```
