---
title: "PubMed API in R"
output:
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "_figures/pubmed/"
)
```

# PubMed API in Python

by Michael T. Moen and Adam M. Nguyen

<div class="rmd-btn-wrapper">
  <a class="rmd-btn"
     href="https://github.com/UA-Libraries-Research-Data-Services/UALIB_ScholarlyAPI_Cookbook/blob/main/rmarkdown/pubmed.Rmd"
     target="_blank"
     rel="noreferrer">
    View RMarkdown File
  </a>
</div>

The PubMed API, part of NCBI's Entrez Programming Utilities (E-utilities), provides programmatic access to biomedical literature from the PubMed database, enabling retrieval of bibliographic data.

Please see the following resources for more information on API usage:
- Documentation
    - <a href="https://www.ncbi.nlm.nih.gov/books/NBK25501/" target="_blank">Entrez Programming Utilities Help</a>
    - <a href="https://www.ncbi.nlm.nih.gov/home/develop/api/" target="_blank">NCBI API Documentation</a>
- Terms
    - <a href="https://www.ncbi.nlm.nih.gov/home/about/policies/" target="_blank">NCBI Terms of Use</a>
    - <a href="https://www.ncbi.nlm.nih.gov/books/NBK25497/" target="_blank">NCBI Entrez Programming Rate Limit</a>
- Data Reuse
    - <a href="https://www.ncbi.nlm.nih.gov/home/about/policies/" target="_blank">NCBI Data Reuse Policy</a>

_**NOTE:**_ The PubMed API (Entrez E-utilities) limits requests to a maximum of 3 requests per second without an API key, and up to 10 requests per second with an API key.

*These recipe examples were tested on March 24, 2026.*

## Setup

The following packages need to be installed into your environment to run the code examples in this tutorial. These packages can be installed with `install.packages()`.

- <a href="https://cran.r-project.org/web/packages/httr/index.html" target="_blank">httr: Tools for Working with URLs and HTTP</a>
- <a href="https://cran.r-project.org/web/packages/jsonlite/index.html" target="_blank">jsonlite: A Simple and Robust JSON Parser and Generator for R</a>

We load the libraries used in this tutorial below:

```{r}
library(httr)
library(jsonlite)
```

## 1. Retieve the Metadata of an Article by PubMed ID

The article we are requesting has a PubMed ID (`pmid`) of 27933103. The `retmode` parameter specifies the file format, which we specify as JSON in this example.

```{r}
ESUMMARY_URL <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esummary.fcgi"
pmid <- "27933103"

params <- list(
  db = "pubmed",
  id = pmid,
  retmode = "json"
)

response <- GET(ESUMMARY_URL, query = params)

# Parse JSON
data <- content(response, as = "text", encoding = "UTF-8")
data <- fromJSON(data)

# Print the field names of the JSON response for the PubMed ID search
names(data$result[[pmid]])
```

Below, we extract from specific information from the response:

```{r}
# Extract the title of the article
data$result[[pmid]]$title
```

```{r}
# Extract author data
data$result[[pmid]]$authors
```

## 2. Retrieve Metadata for a List of PubMed IDs

First, create a list of PubMed IDs:

```{r}
pmids <- c("34813985", "34813932", "34813684", "34813661", "34813372", "34813140", "34813072")
```

Now we can retrieve the data from PubMed and save the data to the list `multi_papers`:

```{r}
multi_papers <- list()
for (pmid in pmids) {

  params <- list(
    db = "pubmed",
    id = pmid,
    retmode = "json"
  )

  Sys.sleep(1)   # Delay between API calls
  response <- GET(ESUMMARY_URL, query = params)

  data <- content(response, as = "text", encoding = "UTF-8")
  data_json <- fromJSON(data)

  # Store paper result using the pmid as the list key
  multi_papers[[pmid]] <- data_json$result[[pmid]]
}

names(multi_papers)
```

Below, we compile a data frame from some of the results stored in the response data.

```{r}
# Compile some of the data of the results to a new data frame
results_df <- data.frame(
  pubdate = sapply(pmids, function(pmid) multi_papers[[pmid]]$pubdate),
  issn = sapply(pmids, function(pmid) multi_papers[[pmid]]$issn),
  source = sapply(pmids, function(pmid) multi_papers[[pmid]]$source),
  stringsAsFactors = FALSE
)

results_df
```

## 3. PubMed API Calls with Requests & Parameters

When searching through the articles, we are given a few of ways of filtering the data. A list of all the available parameters for these requests can be found in the <a href="https://www.ncbi.nlm.nih.gov/books/NBK25499/" target="_blank">NCBI documentation</a>.

In this example, we use the `db` and `term` parameters:
- `db` is set to `pubmed`, which is the database we are working with
- `term` is set to our search query

```{r}
ESEARCH_URL <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi"
params <- list(
  db = "pubmed",
  term = "neuroscience intervention learning",
  retmode = "json"
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Print the number of articles found by the query
data$esearchresult$count
```

```{r}
# Print the number of articles returned by the query
length(data$esearchresult$idlist)
```

```{r}
# Print the PubMed IDs of the first 5 results
data$esearchresult$idlist[1:5]
```

The number of returned IDs can be adjusted with the `retmax` paramater:

```{r}
params <- list(
  db = "pubmed",
  term = "neuroscience intervention learning",
  retmode = "json",
  retmax = 25
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Note how the number of returned articles is now 25 instead of 20
length(data$esearchresult$idlist)
```

The PubMed API can also be used the query to search for an author. To do so, add `[au]` after the name to specify it is an author.

```{r}
params <- list(
  db = "pubmed",
  term = "Darwin[au]",
  retmode = "json"
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Print total number of results of the query
data$esearchresult$count
```

### Sorting Results

We can use the following parameters to store the data for it to be sorted in the same API call:
- `usehistory` is set to `y`, which indicates "yes"
- `sort` is set to `pub date`, which indicates that we are sorting by publication date

```{r}
params <- list(
  db = "pubmed",
  term = "Coral Reefs",
  retmode = "json",
  usehistory = "y",
  sort = "pub date"
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Print the PubMed IDs of the first 5 results
data$esearchresult$idlist[1:5]
```

```{r}
# Compare to unsorted
params <- list(
  db = "pubmed",
  term = "Coral Reefs",
  retmode = "json"
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Notice that the IDs returned are different
data$esearchresult$idlist[1:5]
```

### Searching Based on Publication Type

We can sort by publication type by adding `AND` into the search:
- `AND` in the `term` parameter indicates that both conditions must be present
- `[pt]` specifies that the filter type is publication type

More filters can be found at <a href="https://pubmed.ncbi.nlm.nih.gov/help/" target="_blank">PubMed Help</a>.

```{r}
params <- list(
  db = "pubmed",
  term = "stem cells AND clinical trial[pt]",
  retmode = "json"
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

# Print number of total results
data$esearchresult$count
```

## 4. Visualize Publication Frequencies for Different Topics

### Frequency of Topic sortpubdate Field

Extracting the sortpubdate field for a “hydrogel drug” search results, limited to the "clinical trial" publication type:

```{r}
params <- list(
  db = "pubmed",
  term = "hydrogel drug AND clinical trial[pt]",
  retmode = "json",
  usehistory = "y",
  sort = "pub date",
  retmax = 500
)

Sys.sleep(1)
response <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
ids <- data$esearchresult$idlist

# Print the number of results
length(ids)
```

```{r}
# Print the first 5 results
ids[1:5]
```

```{r}
# Loop through each ID and get the sortpubdate field
# Note that the sortpubdate field may not be equivalent to the publication date
pub_dates <- character(length(ids))
for (i in seq_along(ids)) {
  params <- list(
    db = "pubmed",
    id = ids[i],
    retmode = "json"
  )
  Sys.sleep(1)
  response <- GET(ESUMMARY_URL, query = params)
  data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

  pub_dates[i] <- substr(data$result[[ids[i]]]$sortpubdate, start = 1, stop = 10)
}

# Print the first 5 publication dates
pub_dates[1:5]
```

```{r publication-histogram}
# Convert pub_dates to the Date class
pub_dates <- as.Date(pub_dates)

# Extract years from Date objects
pub_years <- as.integer(format(pub_dates, "%Y"))

# Drop current year from the data
pub_years <- pub_years[pub_years != max(pub_years)]

# Plot data as a histogram
hist(
  pub_years,
  breaks = seq(min(pub_years), max(pub_years), by = 5),
  main = "\"Clinical Trial\" Publications with \"Hydrogel Drug\" in Search Data",
  xlab = "Publication Year",
  ylab = "Number of Publications",
  
  col = "steelblue",
  border = "white"
)
```

### Frequency of Publication for an Author Search

```{r}
params <- list(
  db = "pubmed",
  term = "Reed LK[au]",
  retmode = "json",
  usehistory = "y",
  sort = "pub date",
  retmax = 500
)

Sys.sleep(1)
response  <- GET(ESEARCH_URL, query = params)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
ids <- data$esearchresult$idlist

# Print the number of results
length(ids)
```

```{r}
auth_dates <- character(length(ids))
for (i in seq_along(ids)) {
  params <- list(
    db = "pubmed",
    id = ids[i],
    retmode = "json"
  )
  Sys.sleep(1)
  response <- GET(ESUMMARY_URL, query = params)
  data <- fromJSON(content(response, as = "text", encoding = "UTF-8"))

  auth_dates[i] <- substr(data$result[[ids[i]]]$sortpubdate, start = 1, stop = 10)
}

# Print the first 5 publication dates
auth_dates[1:5]
```

```{r reed-histogram}
auth_dates <- as.Date(pub_dates)
auth_years <- as.integer(format(pub_dates, "%Y"))
auth_years <- pub_years[pub_years != max(pub_years, na.rm = TRUE)]
hist(
  auth_years,
  breaks = seq(min(auth_years), max(auth_years), by = 1),
  main = "Publications by LK Reed",
  xlab = "Publication Year",
  ylab = "Number of Publications",
  
  col = "steelblue",
  border = "white"
)
```
