---

editor_options: 
  markdown: 
    wrap: 72
---

# Scopus API in Python

By Nick Peinitz, Vincent F. Scalfani, and Avery Fernandez

The Scopus API, provided by Elsevier, offers programmatic access to a comprehensive database of abstracts and citations from peer-reviewed literature. It supports advanced search capabilities, author and affiliation retrieval, and citation analysis, facilitating a wide range of academic and research applications.

*This tutorial content is intended to help facilitate academic research.*

Please see the following resources for more information on API usage: - Documentation - <a href="http://www.scopus.com" target="_blank">Scopus</a> - <a href="https://dev.elsevier.com/scopus.html" target="_blank">Scopus API</a> - <a href="https://dev.elsevier.com/api_docs.html" target="_blank">Elsevier API Documentation</a> - <a href="https://dev.elsevier.com/use_cases.html" target="_blank">Elsevier API Use Cases</a> - Terms - <a href="https://dev.elsevier.com/api_service_agreement.html" target="_blank">Elsevier API Service Agreement</a> - <a href="https://dev.elsevier.com/policy.html" target="_blank">Elsevier API Policy</a> - Data Reuse - <a href="https://www.elsevier.com/about/policies-and-standards/research-data" target="_blank">Elsevier Research Data Policy</a> - Scopus Platform

***NOTE:*** Please see access details and rate limit requests for this API in the official documentation.

*These recipe examples were tested on August 4, 2026.*

## Setup

### Import Libraries

The following external libraries need to be installed into your environment to run the code examples in this tutorial:

- <a href="https://github.com/r-lib/httr2" target="_blank">httr2</a>
- <a href="https://github.com/gaborcsardi/dotenv" target="_blank">dotenv</a>
- <a href="https://github.com/r-lib/xml2" target="_blank">xml2</a>
- <a href="https://github.com/tidyverse/dplyr" target="_blank">dplyr</a>
- <a href="https://github.com/tidyverse/purrr" target="_blank">purrr</a>
- <a href="https://github.com/tidyverse/tibble" target="_blank">tibble</a>

We import the libraries used in this tutorial below:

```{r}
library(httr2)
library(dotenv)
library(xml2)
library(dplyr)
library(purrr)
library(tibble)
```

### Import API Key

An API key is required to access the Scopus API. You can sign up for one at the <a href="https://dev.elsevier.com/apikey/manage" target="_blank">Scopus Developer Portal</a>.

We keep our API key in a separate file, a `.env` file, and use the `dotenv` library to access it. If you use this method, create a file named `.env` in the same directory as this notebook and add the following line to it:

``` text
SCOPUS_API_KEY=PUT_YOUR_API_KEY_HERE
```

```{r}
load_dot_env()
API_KEY <- Sys.getenv("SCOPUS_API_KEY")
if (API_KEY == "") {
    message("API key not found. Please set 'SCOPUS_API_KEY' in your .env file.")
} else {
  print("Environment and API key successfully loaded.")
}
```


## 1. Get Author Data

### Number of Records for Author

```{r}
BASE_URL <- "https://api.elsevier.com/content/search/scopus"
query <- "AU-ID(55764087400)"
httpAccept <- "application/json"
data <- tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
  

          resp_body_json(response)

}, error = function(e) {
  message("An error ocurred: ", e$message)
  NULL
})
```

We can take a closer look at the data given back:

```{r}
str(data, max.level = 1)
```

It seems the data is wrapped in a list with the key `search-results`.

```{r}
str(data[["search-results"]], max.level = 1)
```

Inside the `search-results` dictionary, there are six keys: \* `entry` - the actual data we want \* `link` - a link to the API endpoint \* `opensearch:Query` - the query we used to get the data \* `opensearch:itemsPerPage` - the number of items per page \* `opensearch:startIndex` - the starting index of the items \* `opensearch:totalResults` - the total number of results

```{r}
str(data[["search-results"]]$entry, max.level = 1)
```

The `entry` key contains a list of data points, we can see that the first item in the list like so:

```{r}
str(data[["search-results"]]$entry[[1]], max.level = 1)
```

The list inside the `entry` list has information for the individual articles. We can load this into a dataframe to make it easier to work with.

```{r}
entries <- data[["search-results"]][["entry"]]
df <- purrr::map_dfr(entries, function(x) {
  
  x[sapply(x, is.null)] <- NA
  
  x[sapply(x, is.list)] <- lapply(
    x[sapply(x, is.list)],
    function(z) as.character(jsonlite::toJSON(z, auto_unbox = TRUE))
  )
  
  as_tibble_row(x)
})

head(df)
```

```{r}
# See the columns of the DataFrame
colnames(df)
```

```{r}
# Number of rows
nrow(df)
```

```{r}
# We can index data from our new dataframe, df.
# For example, create a vector of just the DOIs
dois <- df[["prism:doi"]]
dois
```

```{r}
# Get a vector of article titles
titles = df[["dc:title"]]
titles
```

```{r}
 # Now a vector of the cited by count
cited_by = df[["citedby-count"]]
cited_by
```

```{r}
# Get sum of cited_by
sum(as.numeric(cited_by))
```


## 2. Get Author Data in a Loop

### Number of Records for Author

```{r}
# Create text file to store author names and their Scopus AUIDs
writeLines(
  c(
    "Emy Decker\t36660678600",
    "Lindsey Lowry\t57210944451",
    "Karen Chapman\t35783926100",
    "Kevin Walker\t56133961300",
    "Sara Whitver\t57194760730"
  ),
"author.txt"
)
```

```{r}
# Load a list of author names and Scopus AUIDs
author_df <- read.delim(
  "author.txt",
  header = FALSE,
  sep = "\t"
)

author_list <- split(author_df, seq_len(nrow(author_df)))
str(author_list)
```

```{r}
# Get number of Scopus records for each author
num_records <- list()

for (i in seq_along(author_list))
{
  author <- author_list[[i]][1, 1]
  authorID <- author_list[[i]][1, 2]

  query <- paste0("AU-ID(", authorID, ")")
  httpAccept <- "application/json"
  tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
          Sys.sleep(1)
          
          data <- resp_body_json(response)

          number_of_records <- data[["search-results"]][["opensearch:totalResults"]]
          num_records[[length(num_records) + 1]] <- list(
              author = author,
              authorID = authorID,
              number_of_records = number_of_records)
  }, error = function(e) {
  message("An error ocurred: ", e$message)
  num_records[[length(num_records) + 1]] <<- list(
      author = author,
      authorID = authorID,
      number_of_records = NA)
})
}
```

```{r}
num_records
```


### Download Record Data

```{r}
# Let's say we want the DOIs and cited by counts in a list
cites <- list()

for (i in seq_along(author_list))
{
  author <- author_list[[i]][1, 1]
  authorID <- author_list[[i]][1, 2]
  
  query <- paste0("AU-ID(", authorID, ")")
  httpAccept <- "application/json"

  tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
          Sys.sleep(1)

          data <- resp_body_json(response)
          
          entries <- data[["search-results"]][["entry"]]
          author_df <- purrr::map_dfr(entries, function(x) {
            
            x[sapply(x, is.null)] <- NA
            
            x[sapply(x, is.list)] <- lapply(
              x[sapply(x, is.list)],
              function(z) as.character(jsonlite::toJSON(z, auto_unbox = TRUE))
            )
            
            as_tibble_row(x)
          })
          
          # Get the DOIs and cited by counts
          dois <- author_df[["prism:doi"]]
          cited_by <- author_df[["citedby-count"]]
          
          # Create a list of lists with author, authorID, DOI, and cited by count
          for (i in seq_along(dois))
          {
            doi <- dois[i]
            cited <- cited_by[i]

            cites[[length(cites) + 1]] <- list(
              author = author,
              authorID = authorID,
              doi = doi,
              cited = cited
            )
          }
  }, error = function(e) {
  message("An error ocurred: ", e$message)
  num_records[[length(num_records) + 1]] <<- list(
      author = author,
      authorID = authorID,
      number_of_records = NA)
  })
}
```

```{r}
# The cites variable is a list of list with the data
# View data for first 5 records
cites[1:5]
```

```{r}
# Add to data frame
cites_df <- bind_rows(cites)
cites_df
```


### Save Record Data to a File

Here is one method if you want to loop over author queries and save all Scopus document data to a file

```{r}
# Load a list of author names and Scopus AUIDs
author_df <- read.delim(
  "author.txt",
  colClasses = "character",
  header = FALSE,
  sep = "\t"
)

author_list <- split(author_df, seq_len(nrow(author_df)))
str(author_list)
```

```{r}
for (i in seq_along(author_list))
{
  authorName <- author_list[[i]][1, 1]
  authorID <- author_list[[i]][1, 2]

  query <- paste0("AU-ID(", authorID, ")")
  httpAccept <- "application/json"

  tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
          Sys.sleep(2)
      
          data <- resp_body_json(response)
      
          # Extract the 'entry' data and convert it to a data frame
          if ("entry" %in% names(data[["search-results"]]))
            df <- bind_rows(data[["search-results"]][["entry"]])
              df[] <- lapply(df, function(col) {
                if (is.list(col)) {
                  sapply(col, jsonlite::toJSON, auto_unbox = TRUE)
                } else {
                    col
                }
              })
      
          # Save to file
          filename <- paste0(
            gsub(" ", "_", authorName),
            "_",
            authorID,
            "_ScopusData.tsv"
          )
          
          write.table(
            df,
            file = filename,
            sep = "\t",
            row.names = FALSE,
            quote = FALSE
          )
          print(paste0("Data for ", authorName, " saved to ", filename))
  }, error = function(e) {
      message("An error ocurred for ", authorName, ": ", e$message)
  })
}
```

```{r}
df_author <- read.delim(
  "Karen_Chapman_35783926100_ScopusData.tsv",
  sep = "\t"
)

head(df_author)
```

```{r}
# Get info about citedby_count
summary(as.numeric(df_author[["citedby.count"]]))
```

```{r}
# Get info about publication titles
x <- df_author[["prism.publicationName"]]

c(
  count <- sum(!is.na(x)),
  unique_vals <- length(unique(x[!is.na(x)])),
  top = names(sort(table(x), decreasing = TRUE))[1],
  freq = max(table(x))
)
```


## 3. Get References via a Title Search

### Number of Title Match Records

```{r}
query <- "TITLE(ChemSpider)"
httpAccept <- "application/json"
tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
        
          data <- resp_body_json(response)
          print(data[["search-results"]][["opensearch:totalResults"]])

}, error = function(e) {
  message("An error ocurred: ", e$message)
  data <- NULL
})
```

```{r}
titleWord_list <- list(
  "ChemSpider", "PubChem", "ChEMBL", "Reaxys", "SciFinder"
)

# Get number of Scopus records for each title search
num_records_title <- list()

for (titleWord in titleWord_list)
{
  query <- paste0("TITLE(", titleWord, ")")
  httpAccept <- "application/json"
  tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()
      
          data <- resp_body_json(response)
          # Extract the total number of results
          numt = data[["search-results"]][["opensearch:totalResults"]]
      
          # Compile saved Scopus data into a list of lists
          num_records_title[[length(num_records_title) + 1]] <- list(
            titleWord,
            numt
          )
      
          # Delay 1 second between API calls to be nice to Elsevier servers
          Sys.sleep(1)
  }, error = function(e) {
    message("An error ocurred: ", e$message)
    num_records_title[[length(num_records_title) + 1]] <- list(
      titleWord,
      numt = NA
    )
  })
}
```

```{r}
num_records_title
```


### Download Title Match Record Data

```{r}
titleWord_list <- list(
  "ChemSpider", "PubChem", "ChEMBL", "Reaxys", "SciFinder"
)

# Get number of Scopus records for each title search
scopus_title_data <- list()

for (titleWord in titleWord_list)
{
  query <- paste0("TITLE(", titleWord, ")")
  httpAccept <- "application/json"
  tryCatch({
          response <- request(BASE_URL) |>
                            req_headers(
                              "X-ELS-APIKey" = API_KEY,
                              "Accept" = httpAccept
                              ) |>
                            req_url_query(
                                query = query
                            ) |>
                            req_perform()

          # Delay 1 second between API calls to be nice to Elsevier servers
          Sys.sleep(1)
      
          data <- resp_body_json(response)
      
          # Extract the 'entry' data and convert it to a DataFrame
          entries <- data[["search-results"]][["entry"]]
          if (!is.null(entries))
          {
            for (entry in entries)
            {
              doi <- if ("prism:doi" %in% names(entry)) entry[["prism:doi"]] else NA
              title <- if ("dc:title" %in% names(entry)) entry[["dc:title"]] else NA
              coverDate <- if ("prism:coverDate" %in% names(entry)) entry[["prism:coverDate"]] else NA
      
          # Append to the list
              scopus_title_data[[length(scopus_title_data) + 1]] <- list(
                titleWord,
                doi,
                title,
                coverDate
              )
            }
          }

  }, error = function(e) {
    message("An error ocurred for ", titleWord, ": ", e$message)
    num_records_title[[length(num_records_title) + 1]] <- list(
      titleWord,
      doi = NA,
      title = NA,
      coverDate = NA
    )
  })
}
```

```{r}
scopus_title_data_df <- bind_rows(
  lapply(scopus_title_data, function(x) {
    tibble(
      titleWord = x[[1]],
      doi = x[[2]],
      title = x[[3]],
      coverDate = x[[4]],
    )
  })
)

scopus_title_data_df
```