---
title: "Springer Nature API in R"
output: 
  html_document:
    keep_md: true
---



# Springer Nature API in R

by Nick Peinitz, Avery Fernandez and Vincent F. Scalfani

These recipe examples use the Springer Nature Open Access API to retrieve metadata and full-text content. The Springer Nature Open Access API includes about 1.5 million full-text records.

*There is also a Full-Text API for subscription content. Please check with your institution for their Text and Data Mining or related License Agreement with Springer Nature.*

Please see the following resources for more information on API usage:
- Documentation
    - <a href="https://dev.springernature.com/" target="_blank">Springer Nature API</a>
    - <a href="https://dev.springernature.com/docs/api-endpoints/open-access/" target="_blank">Springer Nature API Documentation</a>
    - <a href="https://dev.springernature.com/docs/quick-start/api-access/" target="_blank">Springer Nature API Access Quick Start</a>
    - <a href="https://dev.springernature.com/docs/live-documentation/" target="_blank">Springer API Playground</a>
- Terms
    - <a href="https://www.springernature.com/gp/legal/general-terms-of-use/15067848" target="_blank">Springer Nature General Terms of Use</a>
    - <a href="https://dev.springernature.com/terms-conditions" target="_blank">Springer Nature API Terms and Conditions</a>
- Data Reuse
    - <a href="https://www.springernature.com/gp/researchers/text-and-data-mining" target="_blank">Springer Nature Text and Data Mining Policies</a>
    - <a href="https://dev.springernature.com/tdm-reservation-policy/" target="_blank">Springer Nature TDM Reservation Policy</a>


_**NOTE:**_ Check with your institution to determine your <a href="https://dev.springernature.com/docs/rate-limit-details/rate-limits/" target="_blank">API rate limit with Springer Nature</a>.

*If you have copyright or other related text and data mining questions, please contact The University of Alabama Libraries or your respective library/institution.*

*These recipe examples were tested on July 7, 2026.*


## Setup

### Import Libraries

The following external libraries need to be installed into your environment to run the code examples in this tutorial:

- <a href="https://github.com/r-lib/httr2" target="_blank">httr2</a>
- <a href="https://github.com/gaborcsardi/dotenv" target="_blank">dotenv</a>
- <a href="https://github.com/r-lib/xml2" target="_blank">xml2</a>

We import the libraries used in this tutorial below:


```{r}
library(httr2)
library(dotenv)
library(xml2)
```

### Import API Key

Authentication is required to access the Springer Nature API. You can sign up for one at the <a href="https://dev.springernature.com/" target="_blank">Springer Nature Developer Portal</a>.

We keep our API key in a separate file, a `.env` file, and use the `dotenv` library to access it. If you use this method, create a file named `.env` in the same directory as this notebook and add the following line to it:

```text
SPRINGER_API_KEY=PUT_YOUR_API_KEY_HERE
```


```{r}
load_dot_env()
API_KEY <- Sys.getenv("SPRINGER_API_KEY")
if (API_KEY == "") {
    message("API key not found. Please set 'SPRINGER_API_KEY' in your .env file.")
}
```

## 1. Retrieve Full-Text JATS XML of an Article

In this section, we demonstrate how to retrieve the JATS XML content for a specific article based on its DOI.

The <a href="https://jats.nlm.nih.gov/archiving/" target="_blank">JATS</a> XML format is a standard intended for tagging, archiving, and exchanging journal articles. The Springer Nature Open Access API allows us to retrieve articles in JATS XML format.

Key parameters:
* `base_url`: The base URL for the Springer Nature API (Open Access JATS endpoint).
* `q=(doi:DOI)`: The query parameter used to search for an article based on its DOI.
* `api_key`: The query parameter used to pass our valid API key.

More details about the parameters can be found at <a href="https://dev.springernature.com/restfuloperations" target="_blank">Springer Nature Developer Portal</a>.

You can also play around with the API using the <a href="https://dev.springernature.com/docs/live-documentation/" target="_blank">Springer API Playground</a>.

```{r}
# Example article from SpringerOpen Brain Informatics
# This article is under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
# https://doi.org/10.1186/s40708-025-00250-5
doi <- '10.1186/s40708-025-00250-5'

tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/jats") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()

                writeLines(
                    resp_body_string(response),
                    "fulltext.jats"
                )

                message(
                    sprintf(
                        "JATS XML successfully retrieved for DOI %s. Saved to fulltext.jats",
                        doi
                    )
                )
    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JATS XML for DOI %s: %s",
                doi,
                e$message
            )
        )
    }
)
```


## 2. Retrieve Full-Text in a Loop

In many cases, you may have a list of DOIs and want to retrieve the full-text for each of them. Below, we loop over a set of DOIs, retrieve the JATS XML, and store each one in a separate file.
A short delay (`Sys.sleep(1)`) is used to avoid hitting rate limits.

```{r}
dois <- list(
    # Licensed under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
    # https://doi.org/10.1186/s40708-025-00250-5
    '10.1186/s40708-025-00250-5', 
    # Licensed under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
    # https://doi.org/10.1186/s40708-024-00247-6
    '10.1186/s40708-024-00247-6', 
    # Licensed under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
    # https://doi.org/10.1186/s40708-024-00243-w
    '10.1186/s40708-024-00243-w', 
    # Licensed under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
    # https://doi.org/10.1186/s40708-023-00202-x
    '10.1186/s40708-023-00202-x', 
    # Licensed under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
    # https://doi.org/10.1186/s40708-023-00204-9
    '10.1186/s40708-023-00204-9' 
)

for (i in seq_along(dois)) {
  doi <- dois[[i]]
  print(
    paste0(
      "Retrieving JATS XML for DOI ",
      doi,
      "(",
      i,
      "/",
      length(dois),
      ")..."
    )
  )

  tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/jats") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()

                doi_name <- gsub("/", "_", doi, fixed = TRUE)
                doi_name <- gsub('"', "", doi_name, fixed = TRUE)

                output_file = paste0(doi_name, "_jats_text.jats")

                writeLines(
                    resp_body_string(response),
                    output_file
                )

                message(
                        paste0("JATS XML retrieved for DOI",
                        doi,
                        ". Saved to ",
                        output_file,
                        "."
                      )
                )
    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JATS XML for DOI %s: %s",
                doi,
                e$message
            )
        )   
    }
  )

  # Delay to avoid hitting rate limits
  Sys.sleep(1)
}
```


## 3. Acquire and Parse Metadata (JSON)

Alternatively, you can retrieve only the metadata in JSON format by switching the base URL to the `json` endpoint. Then, you can parse relevant fields (e.g., abstract, publication date, etc.).

```{r}
# Example article from SpringerOpen Brain Informatics
# This article is under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
# https://doi.org/10.1186/s40708-025-00250-5
# Harnessing the synergy of statistics and \
# deep learning for BCI competition 4 dataset 4: a novel approach
# Gauttam Jangir, Nisheeth Joshi & Gaurav Purohit 
doi <- '10.1186/s40708-025-00250-5'

metadata_response <- tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/json") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()
    
    resp_body_json(response)

    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JSON metadata for DOI %s: %s",
                doi,
                e$message
            )
        )
        list() 
    }
)

names(metadata_response)
```

Below is an example of how to retrieve specific fields from the metadata, such as the article's **abstract**, **DOI**, **publication date**, **publication name**, and **title**.

```{r}
# metadata_response usually contains the named elements: c("apiMessage", "query", "records")

api_message <- metadata_response[["apiMessage"]]
query_info <- metadata_response[["query"]]
records <- metadata_response[["records"]]
if (is.null(records))
{
  records <- list()
}

print(paste0("API Message:", api_message))
print(paste0("Query:", query_info))

if (length(records) > 0)
{
  # Take the first record if available
  first_record <- records[[1]]
  abstract <- first_record[["abstract"]]
  if (is.null(abstract))
  {
    abstract <- list()
  }
  p <- abstract[["p"]]
  if (is.null(p))
  {
    p <- ""
  }
  cat("Abstract:", p, "\n")

  print(paste0("DOI:", first_record[["doi"]]))
  print(paste0("Online Date:", first_record[["onlineDate"]]))
  print(paste0("Print Date:", first_record[["printDate"]]))
  print(paste0("Publication Name:", first_record[["publicationName"]]))
  print(paste0("Title:", first_record[["title"]]))

  # Get the authors
  creators <- first_record[["creators"]]
  if (is.null(creators))
  {
    creators <- list()
  }

  authors <- c()

  for (author in creators)
  {
    authors <- c(
      authors,
      author[["creator"]]
    )
  }
  print(paste0("Authors:", paste(authors, collapse = ", ")))
} else {
  print("No 'records' were returned in the JSON response.")
}
```

## 4. Parsing XML for Metadata

Sometimes you may want to extract specific pieces of data (e.g., *title*, *abstract*, *authors*, *subjects*) directly from the **JATS XML** instead of the JSON. In this example, we use the R `xml2` package to parse the XML.

The XML structure has a `<records>` tag containing one or more `<article>` tags. Each `<article>` has a `<front>` section for metadata, a `<body>` for main text, and possibly `<back>` for references, etc.

```{r}
# example article from SpringerOpen Brain Informatics
# This article is under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
# https://doi.org/10.1186/s40708-025-00250-5
# Harnessing the synergy of statistics and \
# deep learning for BCI competition 4 dataset 4: a novel approach
# Gauttam Jangir, Nisheeth Joshi & Gaurav Purohit 
doi <- '10.1186/s40708-025-00250-5'

xml_data <- NULL
tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/jats") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()

      xml_data <- resp_body_string(response)

                message(
                    sprintf(
                        "JATS XML successfully retrieved for DOI %s.",
                        doi
                    )
                )
    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JATS XML for DOI %s: %s",
                doi,
                e$message
            )
        )
        NULL
    }
)
```

```{r}
root = NULL

if (!is.null(xml_data))
{
  tryCatch (
    {
      root <- read_xml(xml_data)
      print("XML data successfully parsed.")
    },
    error = function(e) {
        message (
            sprintf (
                "Error parsing XML data: %s",
                e$message
            )
        )   
    }
  )
}
```

```{r}
article_data <- list (
  title = NULL,
  abstract = NULL,
  authors = character(),
  subjects = character()
)

if (!is.null(root))
{
  # Assume there's at least one article under records.
  first_article <- xml_find_first(root, ".//records/article")
  if (!inherits(first_article, "xml_missing"))
  {
    # Title
    title_elem <- xml_find_first(first_article, ".//front/article-meta/title-group/article-title")
    if (!inherits(title_elem, "xml_missing"))
    {
      article_data[["title"]] <- xml_text(title_elem)
    }

    # Abstract
    abstract_elem <- xml_find_first(first_article, ".//front/article-meta/abstract/p")
    if (!inherits(abstract_elem, "xml_missing"))
    {
      article_data[["abstract"]] <- xml_text(abstract_elem)
    }

    # Authors
    authors <- xml_find_all(first_article, ".//front/article-meta/contrib-group/contrib/name")
    for (author in authors)
    {
      # Each author element may have multiple child tags (given, surname, etc.).
      full_name <- paste(xml_text(xml_children(author)), collapse = " ")

      article_data$authors <- c(
        article_data$authors,
        full_name
      )
    }

    # Subjects (keywords)
    subjects <- xml_find_all(first_article, ".//front/article-meta/kwd-group/kwd")
    for (subject in subjects)
    {
      article_data$subjects <- c(
        article_data$subjects,
        xml_text(subject)
      )
    }
  } else {
    print("No article data found in the XML.")
  }
} else {
  print("No valid XML data to parse.")
}

str(article_data)
```

## 5. Parsing XML for Figure Captions

Figure captions often appear under `<fig>` tags inside the `<body>` element. Each figure may have a `<label>` tag for the figure number and a `<caption>` tag for the figure's description.

```{r}
# Example article from SpringerOpen Brain Informatics
# This article is under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
# https://doi.org/10.1186/s40708-025-00250-5
# Harnessing the synergy of statistics and \
# deep learning for BCI competition 4 dataset 4: a novel approach
# Gauttam Jangir, Nisheeth Joshi & Gaurav Purohit 
doi <- '10.1186/s40708-025-00250-5'

xml_data <- NULL
tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/jats") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()

      xml_data <- resp_body_string(response)

                message(
                    sprintf(
                        "JATS XML successfully retrieved for DOI %s.",
                        doi
                    )
                )
    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JATS XML for DOI %s: %s",
                doi,
                e$message
            )
        )   
    }
)
```


```{r}
root = NULL

if (!is.null(xml_data))
{
  tryCatch (
    {
      root <- read_xml(xml_data)
      print("XML data successfully parsed.")
    },
    error = function(e) {
        message (
            sprintf (
                "Error parsing XML data: %s",
                e$message
            )
        )   
    }
  )
} else {
  print("No valid XML data to parse.")
}
```


```{r}
figures_data <- list()

if (!inherits(root, "xml_missing") && !is.null(root))
{
  # Find all <fig> elements within the <body> of the XML
  figures <- xml_find_all(root, ".//body//fig")
  for (fig in figures)
  {
    # Extract the <label> element (e.g., "Figure 1") if it exists
    label <- xml_find_first(fig, "label")
    # Extract the <caption> element (e.g., description of the figure) if it exists
    caption <- xml_find_first(fig, "caption")

    # Get the text content of the label, or use an empty string if not present
    if (!inherits(label, "xml_missing"))
    {
      label_text <- xml_text(label)
    } else {
      label_text <- ""
    }

    # Get the text content of the caption, joining all inner text, or use an empty string if not present
    if(!inherits(caption, "xml_missing"))
    {
      caption_text <- xml_text(caption)
    } else {
      caption_text <- ""
    }

    # Append the figure's label and caption as a named list
    figures_data[[length(figures_data) + 1]] <- list(
      label = label_text,
      caption = trimws(caption_text)
    )
  }
} else {
  # If the XML root is not valid, print an error message
  print("No valid XML data to parse.")
}

# Check if any figures were found and processed
if (length(figures_data) > 0)
{
  print("Figures data:")
  # Iterate through the collected figures and print their details
  for (i in seq_along(figures_data))
  {
    fig_data <- figures_data[[i]]
    print(paste0("Figure ", i, ":"))
    print(paste0("Label: ", fig_data[["label"]]))
    print(paste0("Caption: ", fig_data[["caption"]]))
    print("\n")
  }
} else {
  # If no figures were found, print a message indicating this
  print("No figures data found in the XML.")
}
```

## 6. Extracting Full-Text from the Body

Finally, we can extract a rough "plain text" version of the article body by iterating through each element (e.g., `<sec>`, `<p>`), capturing the text, and joining it into a single string. This can help with quick text-based analyses.

```{r}
# Example article from SpringerOpen Brain Informatics
# This article is under CC-BY-4.0 license https://creativecommons.org/licenses/by/4.0/
# https://doi.org/10.1186/s40708-025-00250-5
# Harnessing the synergy of statistics and \
# deep learning for BCI competition 4 dataset 4: a novel approach
# Gauttam Jangir, Nisheeth Joshi & Gaurav Purohit 
doi <- '10.1186/s40708-025-00250-5' 

xml_data <- NULL
tryCatch (
    {
    response <- request("https://api.springernature.com/openaccess/jats") |>
                    req_url_query(
                        q = sprintf('doi:"%s")', doi),
                        api_key = API_KEY
                ) |>
                req_perform()

      xml_data <- resp_body_string(response)

                message(
                    sprintf(
                        "JATS XML successfully retrieved for DOI %s.",
                        doi
                    )
                )
    },
    error = function(e) {
        message (
            sprintf (
                "Error retrieving JATS XML for DOI %s: %s",
                doi,
                e$message
            )
        )   
    }
)
```


```{r}
root = NULL

if (!is.null(xml_data))
{
  tryCatch (
    {
      root <- read_xml(xml_data)
      print("XML data successfully parsed.")
    },
    error = function(e) {
        message (
            sprintf (
                "Error parsing XML data: %s",
                e$message
            )
        )   
    }
  )
} else {
  print("No valid XML data to parse.")
}
```


```{r}
full_text <- ""
if (!is.null(root) && !inherits(root, "xml_missing"))
{
  # Find the body element
  body <- xml_find_first(root, ".//body")
  if (!is.null(body) && !inherits(body, "xml_missing"))
  {
    # We'll store text in a list, then join them.
    text_parts <- list()

    # We can iterate over each top-level child in the body.
    # Typically <sec> tags hold paragraphs, etc.
    sections <- xml_children(body)
    for (section in sections)
    {
      # For each subsection, gather all text.
      sub_sections <- xml_children(section)
      for (sub_section in sub_sections)
      {
        text_parts <- c(
          text_parts,
          xml_text(sub_section)
        )
      }
      # Combine everything
      full_text <- paste(text_parts, collapse = "\n")
      full_text <- trimws(full_text)
    }
  } else {
    print("No body content found in the XML.")
  }
} else {
  print("No valid XML data to parse.")
}

if (length(full_text) > 0 && full_text != "")
{
  writeLines(
    full_text,
    "fulltext.txt"
    )
  print("Full text extracted and saved to 'fulltext.txt'.")  
} else {
  print("No full text content found in the XML.")
}
```


```{r}
# Output a portion of the full text to the console
print(substr(full_text, 1, 395))
```
