---
title: "U.S. Census Data API in R"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "_figures/us-census/"
)
```

# U.S. Census Data API in R

by Michael T. Moen and Adam M. Nguyen

<div class="rmd-btn-wrapper">
  <a class="rmd-btn"
     href="https://github.com/UA-Libraries-Research-Data-Services/UALIB_ScholarlyAPI_Cookbook/blob/main/rmarkdown/us-census.Rmd"
     target="_blank"
     rel="noreferrer">
    View RMarkdown File
  </a>
</div>

The U.S. Census Data API provides programmatic access to demographic, economic, and geographic data collected by the U.S. Census Bureau. It enables users to retrieve and analyze a wide variety of data sets, including Census surveys and population statistics.

Please see the following resources for more information on API usage:
- Documentation
    - <a href="https://www.census.gov/data/developers/data-sets.html" target="_blank">U.S. Census Bureau API Datasets</a>
    - <a href="https://www.census.gov/data/developers/guidance/api-user-guide.html" target="_blank">U.S. Census Bureau API User Guide</a>
    - <a href="https://www.census.gov/data/developers/about.html" target="_blank">U.S. Census Bureau API Overview</a>
    - <a href="https://api.census.gov/data.html" target="_blank">U.S. Census Data Discovery Tool</a>
- Terms
    - <a href="https://www.census.gov/data/developers/about/terms-of-service.html" target="_blank">U.S. Census Bureau API Terms of Service</a>
- Data Reuse
    - <a href="https://www.census.gov/about/policies/open-gov/open-data.html" target="_blank">U.S. Census Bureau Open Data Policy</a>

**_NOTE:_** Please see access details and rate limit requests for this API in the official documentation.

*These recipe examples were tested on March 23, 2026.*

## Setup   

### Load Libraries

The following packages need to be installed into your environment to run the code examples in this tutorial. These packages can be installed with `install.packages()`.

- <a href="https://cran.r-project.org/web/packages/httr/index.html" target="_blank">httr: Tools for Working with URLs and HTTP</a>
- <a href="https://cran.r-project.org/web/packages/jsonlite/index.html" target="_blank">jsonlite: A Simple and Robust JSON Parser and Generator for R</a>

We load the libraries used in this tutorial below:

```{r import-libraries}
library(httr)
library(jsonlite)
```

### Import API Key

An API key is required to access the U.S. Census Data API. You can sign up for one at the <a href="https://api.census.gov/data/key_signup.html" target="_blank">Key Signup page</a>.

We keep our token in a `.Renviron` file that is stored in the working directory and use `Sys.getenv()` to access it. The `.Renviron` should have an entry like the one below.

```text
CENSUS_API_KEY="PUT_YOUR_API_KEY_HERE"
```

Below, we can test to whether the key was successfully imported.

```{r import-api-key}
if (nzchar(Sys.getenv("CENSUS_API_KEY"))) {
  print("API key successfully loaded.")
} else {
  warning("API key not found or is empty.")
}
```

## 1. Get Population Estimates of Counties by State

*Note: This data includes the District of Columbia and Puerto Rico*

For obtaining data from the Census API, it is helpful to first obtain a list of state IDs:

```{r}
# Set the base URL that will be used throughout this tutorial
BASE_URL <- "https://api.census.gov/data/"

# The parameters specify what data we want to retrieve
params <- list(
  get = "NAME",
  `for` = "state:*",   # This will grab the names of all states in the US
  key = Sys.getenv("CENSUS_API_KEY")
)
year <- 2019

# Make the request to the Census API
response <- GET(paste0(BASE_URL, year, "/pep/population"), query = params)

# Get the JSON data from the response
states <- data.frame(fromJSON(rawToChar(response$content)))

# Print first 6 rows of the resulting data frame
head(states)
```

```{r}
# Add rownames and drop the first row
names(states) <- c("state", "fips_state")
states <- states[-1, ]
head(states)
```

```{r}
params <- list(
  get = "NAME,POP",
  `for` = "county:*",
  key = Sys.getenv("CENSUS_API_KEY")
)
year <- 2019

response <- GET(paste0(BASE_URL, year, "/pep/population"), query = params)
raw_df <- data.frame(fromJSON(rawToChar(response$content)))
names(raw_df) <- raw_df[1, ]
raw_df <- raw_df[-1, ]

df <- data.frame(
  # Merge State and County FIPS codes
  fips = paste0(raw_df$state, raw_df$county),
  # Split County and State name into different columns
  county = sub(",.*$", "", raw_df$NAME),
  state = sub("^[^,]*,\\s*", "", raw_df$NAME),
  pop_2019 = as.integer(raw_df$POP)
)

# Print number of rows in the data frame (one per U.S. county)
nrow(df)
```

```{r}
# Print first 6 rows of the result data frame
head(df)
```

```{r}
# Filter data frame by state
head(df[df$state == "Missouri", ])
```

```{r}
# Filter counties over a certain population threshold
df[df$pop_2019 > 3000000, ]
```

## 2. Get Population Estimates Over a Range of Years

We can use similar code as before, but now loop through different population estimate datasets by year. Here are the specific endpoints used:

- <a href="https://api.census.gov/data/2015/pep/population/examples.html" target="_blank">Vintage 2015 Population Estimates</a>
- <a href="https://api.census.gov/data/2016/pep/population/examples.html" target="_blank">Vintage 2016 Population Estimates</a>
- <a href="https://api.census.gov/data/2017/pep/population/examples.html" target="_blank">Vintage 2017 Population Estimates</a>
- <a href="https://api.census.gov/data/2018/pep/population/examples.html" target="_blank">Vintage 2018 Population Estimates</a>

```{r}
params <- list(
  get = "GEONAME,POP",
  `for` = "county:*",
  key = Sys.getenv("CENSUS_API_KEY")
)

for (year in 2015:2018) {
  response <- GET(paste0(BASE_URL, year, "/pep/population"), query = params)
  Sys.sleep(1)
  
  # Process data from the response
  raw_df <- data.frame(fromJSON(rawToChar(response$content)))
  names(raw_df) <- raw_df[1, ]
  raw_df <- raw_df[-1, ]
  
  raw_df <- data.frame(
    # Merge State and County FIPS codes
    fips = paste0(raw_df$state, raw_df$county),
    # Split County and State name into different columns
    pop = as.integer(raw_df$POP)
  )
  
  # Rename the pop column to contain the year
  names(raw_df)[names(raw_df) == "pop"] <- paste0("pop_", year)
  
  # Merge the new data with the overall data frame
  df <- merge(df, raw_df, by = "fips")
}

# Reorder columns
df <- df[, c("state", "county", "fips", "pop_2015", "pop_2016", "pop_2017",
             "pop_2018", "pop_2019")]

# Print updated data frame
head(df)
```

## 3. Plot Population Change

This data is based off the <a href="https://api.census.gov/data/2021/pep/population/examples.html" target="_blank">2021 Population Estimates</a> dataset.

The percentage change in population is from July 1, 2020 to July 1, 2021 for states (including the District of Columbia and Puerto Rico).

```{r}
params <- list(
  get = "NAME,POP_2021,PPOPCHG_2021",
  `for` = "state:*",
  key = Sys.getenv("CENSUS_API_KEY")
)
year <- 2021

response <- GET(paste0(BASE_URL, year, "/pep/population"), query = params)
data <- data.frame(fromJSON(rawToChar(response$content)))
data <- data[-1, ]

# Print number of results
nrow(data)
```

```{r}
# Rename columns
names(data) <- c("state", "pop_2021", "percent_pop_change_2021", "fips_state")

# Sort data frame alphabetically by state
data <- data[order(data$state), ]

# Convert state to a factor for plotting
data$state <- factor(data$state, levels = rev(sort(data$state)))

# Print first 6 states
head(data)
```

```{r population-plot, fig.width = 6, fig.height = 8}
# Expand left margin
par(mar = c(3, 7, 3, 3))

# Make a scatter plot
plot(
  data$percent_pop_change_2021,
  data$state,
  pch = 21,
  bg = adjustcolor("#1f77b4", 0.7),
  col = "white",
  cex = 1.2,
  xlab = "% Population Change",
  ylab = "",
  main = "Population Change from 2020 to 2021",
  frame.plot = FALSE,
  yaxt = "n"
)

# Add line at x = 0
abline(v = 0, col = "gray", lty = 2)

# Add state labels along y-axis
axis(
  side = 2,
  at = seq_along(levels(data$state)),
  labels = levels(data$state),
  las = 2,
  cex.axis = 0.7
)
```
