---
title: "World Bank API in R"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "_figures/world-bank/"
)
```

# World Bank API in R

by Michael T. Moen, Vishank Patel, and Adam M. Nguyen

The World Bank offers a suite of APIs providing access to a vast array of global development data, including economic indicators, project information, and more. These APIs enable users to programmatically retrieve data for analysis, application development, and research purposes.

Please see the following resources for more information on API usage:
- Documentation
    - <a href="https://datahelpdesk.worldbank.org/knowledgebase/articles/889392-about-the-indicators-api-documentation" target="_blank">World Bank Indicators API Documentation</a>
    - <a href="https://datahelpdesk.worldbank.org/knowledgebase/articles/1886698-data-catalog-api" target="_blank">World Bank Data Catalog API Documentation</a>
    - <a href="https://data.worldbank.org/" target="_blank">World Bank Data Catalog</a>
    - <a href="https://datahelpdesk.worldbank.org/knowledgebase/articles/902064-development-best-practices" target="_blank">World Bank Development Best Practices</a>
- Terms
    - <a href="https://www.worldbank.org/en/about/legal/terms-and-conditions#:~:text=You%20may%20not%20use%20the,your%20use%20of%20the%20APIs." target="_blank">World Bank Terms of Use</a>
    - <a href="https://data.worldbank.org/summary-terms-of-use" target="_blank">World Bank Summary Terms of Use</a>
- Data Reuse
    - <a href="https://www.worldbank.org/en/about/legal/terms-of-use-for-datasets" target="_blank">World Bank Data Licensing and Terms of Use</a>

*These recipe examples were tested on April 2, 2026.*

## Setup

### Import Libraries

The following external libraries need to be installed into your environment to run the code examples in this tutorial:

- <a href="https://cran.r-project.org/web/packages/countrycode/index.html" target="_blank">countrycode: Convert Country Names and Country Codes</a>
- <a href="https://cran.r-project.org/web/packages/httr/index.html" target="_blank">httr: Tools for Working with URLs and HTTP</a>
- <a href="https://cran.r-project.org/web/packages/jsonlite/index.html" target="_blank">jsonlite: A Simple and Robust JSON Parser and Generator for R</a>
- <a href="https://cran.r-project.org/web/packages/plotly/index.html" target="_blank">plotly: Create Interactive Web Graphics via 'plotly.js'</a>

We import the libraries used in this tutorial below:

```{r, message=FALSE}
library(countrycode)
library(httr)
library(jsonlite)
library(plotly)
```

### DataBank Bulk Download

The World Bank offers bulk downloads of specific datasets from <a href="https://databank.worldbank.org/home.aspx" target="_blank">DataBank Bulk Download</a> or <a href="https://microdata.worldbank.org/index.php/home" target="_blank">DataBank Bulk Microdata Download</a>. If you require extensive or complete data, consider using these bulk downloads. For smaller requests or interactive analysis, the standard API calls work well.

## 1. Comparing GDP per Capita by Country

In this example, we look at the GDP per capita of each country using the <a href="https://data.worldbank.org/indicator/NY.GDP.PCAP.CD" target="_blank">NY.GDP.PCAP.CD</a> indicator. Note we examine both countries and dependent territories defined as having <a href="https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3" target="_blank">ISO 3166-1 alpha-3</a> (also known here as ISO 3) codes.

```{r}
BASE_URL <- "http://api.worldbank.org/v2/country/all/indicator/"
indicator <- "NY.GDP.PCAP.CD"    # Indicator code for GDP per capita (current USD)

params <- list(
  date = 2024,
  format = "json",
  per_page = 500
)
response <- GET(paste0(BASE_URL, indicator), query = params)

# Status code 200 indicates success
response$status_code
```

```{r}
# Extract JSON data from response
df <- fromJSON(rawToChar(response$content))[[2]]

# Drop unneeded columns
df <- df[, !(names(df) %in% c("country", "indicator", "unit", "obs_status", "decimal"))]

# Print first result
head(df)
```

Below, we use the `countrycode` package to retrieve the names of the countries based on their ISO 3 codes. Since the World Bank also maintains listings for different aggregations, we also drop those from the data below:

```{r, warning=FALSE}
# Get country names by ISO 3 codes
df$country_name <- countrycode(
  df$countryiso3code,
  origin = "iso3c",
  destination = "country.name"
)

# Drop rows that don't have a related country name
df <- df[!is.na(df$country_name), ]

# Find and sort top 20
top20 <- head(df[order(-df$value), ], 20)

# Display results
top20
```

```{r top20_gdp_per_capita, fig.width=10, fig.height=6}
# Margins adjusted for rotated labels
par(mar = c(5, 5, 4, 2) + 0.1)

# Create barplot
bp <- barplot(
  top20$value / 1000,
  col = "#4C72B0",
  border = NA,
  names.arg = NA,
  main = "Top 20 Countries by GDP per Capita (2024)",
  xlab = NA,
  ylab = "GDP per Capita (Thousand USD)"
)

# Add rotated bar labels
text(
  x = bp,
  y = par("usr")[3] - 0.04 * diff(par("usr")[3:4]),
  labels = top20$country_name,
  srt = 30,
  adj = 1,
  xpd = TRUE,
  cex = 0.8
)

# Draw axis again so it appears above grid
box(bty = "l", col = "#CCCCCC")
```

## 2. Comparing Scientific and Technical Journal Articles by Country

In this example, we look at the "scientific and technical journal articles" of each country using the <a href="https://data.worldbank.org/indicator/IP.JRN.ARTC.SC" target="_blank">IP.JRN.ARTC.SC</a> indicator and the total population of each country using the <a href="https://data.worldbank.org/indicator/SP.POP.TOTL" target="_blank">SP.POP.TOTL</a>. Note we examine both countries and dependent territories defined as having <a href="https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3" target="_blank">ISO 3166-1 alpha-3</a> (also known here as ISO 3) codes.

```{r, warning=FALSE}
BASE_URL <- "https://api.worldbank.org/v2/country/all/indicator/"
indicators <- c(
  "IP.JRN.ARTC.SC" = "Article Count",
  "SP.POP.TOTL" = "Population"
)
params <- list(
  date = 2021,
  format = "json",
  per_page = 500
)

dfs <- list()

for (indicator in names(indicators)) {
  colname <- indicators[[indicator]]
  response <- GET(paste0(BASE_URL, indicator), query = params)
  response <- fromJSON(rawToChar(response$content))
  df_temp <- response[[2]]
  df_temp <- df_temp[, c("countryiso3code", "value")]
  names(df_temp)[names(df_temp) == "value"] <- colname
  dfs[[indicator]] <- df_temp
}

# Merge both datasets
df <- merge(dfs[[1]], dfs[[2]], by = "countryiso3code")

# Get country names by ISO 3 codes
df$`Country Name` <- countrycode(
  df$countryiso3code,
  origin = "iso3c",
  destination = "country.name"
)

# Remove aggregates like "World", "High income", etc.
df <- df[!is.na(df$`Country Name`), ]

head(df)
```

```{r}
# Calculate article count per million people
df$`Article Count per Million` = df$`Article Count` / df$`Population` * 1000000

head(df)
```

```{r, warning=FALSE}
fig <- plot_ly(
  data = df,
  type = "choropleth",
  locations = ~countryiso3code,
  z = ~`Article Count per Million`,
  text = ~`Country Name`,
  colorscale = "Viridis",
  locationmode = "ISO-3",
  hovertemplate = paste(
    "<b>%{text}</b><br>",
    "Article Count per Million: %{z:.2f}",
    "<extra></extra>"
  ),
  colorbar = list(
    title = list(text = "Article Count per Million")
  )
) %>%
  layout(
    title = list(text = "Technical Journal Articles by Country (2021)"),
    margin = list(l = 50, r = 50, b = 50, t = 50),
    geo = list(
      projection = list(type = "natural earth"),
      showframe = FALSE,
      showcoastlines = TRUE,
      coastlinecolor = "lightgray"
    )
  )

fig
```

```{r, include=FALSE}
# Note: This cell will not run or appear in the built book.
# However, it is included here to show how to save the Plotly figure as an HTML file that can
# be embedded in the book.
htmlwidgets::saveWidget(
  widget = fig,
  file = "../src/_static/plotly/technical_journal_articles_2021_r.html",
  selfcontained = TRUE
)
```

```{raw} html
<iframe
    src="../_static/plotly/technical_journal_articles_2021_r.html"
    width="100%"
    height="500"
    style="border:0;"
    loading="lazy">
</iframe>
```
